\chapter*{The Viable System Model}
\section*{How to design a healthy business: The use of the Viable System Model in the diagnosis and design of organisational structures in co-operatives and other social economy enterprises}
\subsection*{A manual for the diagnosis and design of organisational structures to enable social economy enterprises and function with increased efficiency without compromising democratic principles}

{\footnotesize Based on \textit{The Viable Systems Model Pack}, originally published as part of the \textit{SMSE Strategic Management in the Social Economy} training programme
carried out by ICOM, CRU, CAG and Jon Walker with the financial assistance of Directorate General XXIII of the Commission of the European Communities.

The original version was completed October 1991. This \nth{3} revised version incorporates new material.

This print version was created by Joel Kappes based upon the HTML version constructed by John Waters, who also prepared the diagrams and the bibliography. The latter could be found under \href{https://vsmg.lrc.org.uk}{https://vsmg.lrc.org.uk} as of January 2023.

Copyright © 1991 by ICOM, CRU, CAG and Jon Walker. Copyright © 1998, 2020 by Jon Walker.

Version 3.2 - Last modified 21st August 2020 to add a Creative Commons non-commercial licence since it has been brought to our attention that this copyrighted material has not only been cloned by others (which is not a problem) but also charged for on a commercial basis (which is completely unacceptable). A more completely revised version will be released in due course. }